Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-02 doc/invt.tex
small corrections and edits to invt.tex

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002, laio2008, barducci2011, sutto2012}
are two widely-used techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts an extended window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

@@

In this study,
we shall give a proof of the inverse-time formula,
and try to generalize it to
updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is generally one of the best
updating schemes in terms of convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i
  =
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the correlation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(z)
&\equiv
e^{-\lambda \, q(t) + \lambda \, z}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
&\left\langle x^2(t) \right\rangle
=
\langle x^2(0) \rangle e^{-2 \, \lambda \, q(t)}
\notag
\\
&\qquad
+
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average}
\end{align}

We shall further demand that
%
\begin{equation}
  \lim_{t \to \infty} q(t) \to \infty.
  \label{eq:qt_limit}
\end{equation}
%
Then, for a large $t$,
the second term on the right-hand side
of Eq. \eqref{eq:xt_solution} can be neglected, and
%
\begin{align}
\left\langle x^2(t) \right\rangle
%&=
%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%\notag
%\\
&=
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average_asym}
\end{align}

Variating this expression and
using the Euler-Lagrange equation yields
$$
\begin{aligned}
0
&=
\frac{d}{d\tau} \int_0^t
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt'
\\
&= \int_0^t
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $\tau$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition}
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm, and differentiating this with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

\subsubsection{\label{sec:Gamma}
Integrals of the correlation functions}

@@

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_nn}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_k = 1 - 2 \sin^2 \frac{ (k - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_k$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_k$'s.
%
The correlation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value $\kappa_k(0) = 1$ being the same as
that in perfect sampling.
%
Thus,
$\Gamma_k$ is roughly twice
the autocorrelation time,
and from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

\note{If we can interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}

%
%We can study two limiting cases.
%
%For $\tau \ll t$,
%the dominant contribution comes from
%the term(s) with the smallest eigenvalue $\lambda_k = \lambda_n$,
%and
%$$
%\ddot u_n(\tau) = 0,
%\quad
%\dot u_n(\tau) = c,
%\quad
%\mathrm{and}\;
%u_n(\tau) = c \, (\tau + t_0),
%$$
%which means
%$$
%\alpha(\tau) \approx \frac{1}{\lambda_n \, (\tau + t_0)}.
%$$

\subsubsection{\label{sec:optWL}
Optimality of the single-bin scheme}

We now show
that the single-bin scheme
is an optimal scheme.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different eigenvalues, $\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_k \ge 0$
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t') \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
To see this, we first observe that
$$
\int_0^t
  dt' \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
\ge 0,
$$
holds for any $Y$.
%
So the left-hand side,
as a quadratic polynomial of $Y$,
$A \, Y^2 + B \, Y + C$,
must have a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
This condition gives the desired inequality.
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
expressed in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t') \right) = c_k$
for all $k = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_k \, (t' + t_{k0} ) },
\quad
\mathrm{with\;\;}
t_{k0} = \frac{             t            }
              { e^{ \lambda_k q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t') = c_k \left(t' + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(t) = 1$ and $u_k(0) = e^{-\lambda_k \, q(t)}$,
  we have
  $$
  c_k = \frac{ 1 }{ t + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { t + t_{k0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(t) + \lambda_k \, q(t')
  = \log c_k + \log\left( t' + t_{k0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_k \, \alpha(t') = \frac{ 1 } { t' + t_{k0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is generally the most efficient
updating scheme.

@@

For an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
can lead to an algebraically complex expression for $\alpha(t)$.
%
In practice, it is more useful to find
a simple expression for $\alpha(t)$
that is near optimal.
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

@@

Clearly, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).

@@

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  \mu(x) \, \cos \bigl( (k-1) \, x \bigr) \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi \mu(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

@@

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel}
to sample the state space.
%
In each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
In the first sampling scheme,
we choose $j$ randomly out of the $n$ bins
to emulate the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}.
%
We refer to this scheme as the global sampling scheme.
%
In the second sampling scheme,
we limit $j$ to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme models local sampling processes
typical in MD and straightforward MC simulations.
%
We refer to this as the local sampling scheme,

@@

We first tested the single-bin (WL) updating scheme
using the above protocol.
%
Here, we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling schemes,
the optimal $\lambda$ occurred around $1.0$,
the optimal value in the asymptotic regime.
%
Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
represents the relative updating magnitude.
%
Figure \ref{fig:err_singlebin} also shows that
a smaller than optimal value of $1/\lambda$
tends to increase the error more rapidly
than a larger than optimal value,
suggesting that the latter is the safer choice
when the optimal value is unknown.

Our final observation is that the error function
in the two cases differed only by a multiplicative constant.
%
This again agreed with the theory.
%
Since all eigenvalues of the single-bin scheme were
the same, $\lambda_2 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling scheme only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the multiplicative constant mentioned above.

>>>>

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002, laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts an extended window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

@@

In this study,
we shall give another proof of the inverse-time formula.
%
We then adapt the formula
to updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
We shall show that the optimal updating magnitude
depends nontrivially on the simulation length
and other parameters.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is best in asymptotic convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i
  =
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{-\lambda \, q(t) + \lambda \, q'}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
&\left\langle x^2(t) \right\rangle
=
\langle x^2(0) \, \rangle e^{-2 \, \lambda \, q(t)}
\notag
\\
&\qquad
+
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average}
\end{align}

We now fix the endpoint value of $q(t)$
and variate the curve.
%
This is equivalent to variating $u$
with fixing endpoint values,
$u\bigl( q(0) \bigr)  = e^{-\lambda \, q(t)}$
and
$u\bigl( q(t) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the variation process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{d\tau} \int_0^t
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt'
\\
&= \int_0^t
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $\tau$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition}
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm, and differentiating this with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

\subsubsection{\label{sec:Gamma}
Integrals of the autocorrelation functions}

@@

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_nn}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_k = 1 - 2 \sin^2 \frac{ (k - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_k$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_k$'s.
%
The autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value $\kappa_k(0) = 1$ being the same as
that in perfect sampling.
%
Thus,
$\Gamma_k$ is roughly twice
the autocorrelation time,
and from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}

%\subsubsection{\label{twolimits}
%Two limits of the optimal schedule, $\alpha(t)$}
%
%
%The optimal $\alpha(t)$ that minimizes the error
%Eq. \eqref{eq:error_split}
%is still difficult to find analytically.
%%
%However, we can study two limiting cases,
%for the asymptotic part, Eq. \eqref{eq:error_asym1}.
%
%For $\tau \ll t$,
%the dominant contribution comes from
%the term(s) with the smallest eigenvalue $\lambda_k = \lambda_n$,
%and
%$$
%\dot u_n(\tau) = c,
%\quad
%\mathrm{and}\;
%u_n(\tau) = c \, (\tau + t_0),
%$$
%which means
%$$
%\alpha(\tau) \approx \frac{1}{\lambda_n \, (\tau + t_0)}.
%$$
%
%
%For $\tau \approx t$,
%$u_k(q) \approx 1$,

\subsubsection{\label{sec:optWL}
Asymptotic optimality of the single-bin (WL) scheme}

We now give an interesting deduction
following from the error functional
given in Eq. \eqref{eq:error_asym1}:
the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different eigenvalues, $\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of numbers, $c_k \ge 0$,
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t') \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k q(t) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
This inequality follows from the fact
that the following quadratic polynomial of $Y$,
$$
\int_0^t
  dt' \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
%\ge 0,
$$
is always nonnegative, and hence
has a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
This condition furnishes the desired inequality.
%
Note that the last expression of \eqref{eq:CSineq}
depends on $q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t') \right) = c_k$
for all $k = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
%
$$
\alpha(t') = \frac{              1              }
                  { \lambda_k \, (t' + t_{k0} ) },
\quad
\mathrm{with\;\;}
t_{k0} = \frac{             t            }
              { e^{ \lambda_k q(t) } - 1 }.
$$
%
Such a solution is possible only if
%
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t') = c_k \left(t' + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(t) = 1$ and $u_k(0) = e^{-\lambda_k \, q(t)}$,
  we have
  $$
  c_k = \frac{ 1 }{ t + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { t + t_{k0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(t) + \lambda_k \, q(t')
  = \log c_k + \log\left( t' + t_{k0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_k \, \alpha(t') = \frac{ 1 } { t' + t_{k0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of asymptotic convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is generally the most efficient
updating scheme.

We must, however, emphasize that the above result
holds only under the long time limit,
and it is not necessarily true
for finite length simulations.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
before starting the schedule of, $\alpha(t)$.
%
Then for shorter times, $t \approx 0$,
the error is given by Eq. \eqref{eq:error_eql},
which is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
In practice, real simulations
may have long achieved the desired precision
before reaching the asymptotic regime,
especially with some $\lambda_k$'s close to $0$.

@@

Generally, for an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a simple expression for $\alpha(t)$
that is optimal with respect to some free parameter(s).
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

@@

Clearly, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E \ge E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).
%
For a set of nonidentical $\lambda_k$'s,
the equality is reached only at $t = 0$.

@@

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  \mu(x) \, \cos \bigl[ (k-1) \, x \bigr] \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi \mu(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

@@

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
The difference is that,
in the first sampling scheme,
we choose $j$ randomly out of the $n$ bins
to emulate the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}.
%
We refer to this as the global sampling scheme.
%
In the second scheme,
we limit $j$ to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme models local sampling processes
typical in MD and straightforward MC simulations.
%
We refer to this as the local sampling scheme,

@@

We first tested the single-bin (WL) updating scheme
using the above protocol.
%
Here, we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling schemes,
the optimal $\lambda$ occurred around $1.0$,
the optimal value in the asymptotic regime.
%
Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
represents the relative updating magnitude.
%
Figure \ref{fig:err_singlebin} also shows that
a smaller than optimal value of $1/\lambda$
tends to increase the error more rapidly
than a larger than optimal value,
suggesting that the latter (over-updating)
is generally preferable to the latter (under-updating)
when the optimal value is unknown.

Our final observation is that the error functions
in the two cases shared the same shape,
differing only by a multiplicative constant.
%
This again agreed with the theory.
%
Since all eigenvalues of the single-bin scheme were
the same, $\lambda_2 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling scheme only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.
